Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2020-11-09 manu_draft.Rmd
using argument names in list to pmap function

For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South San Francisco Bay.  This model was chosen among the others because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)). Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)). All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record. Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).

@@

Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December (Figure \@ref(fig:trnddat)).  This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.  Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 57.7% increase, although the change is insignificant.  However, the change from 2000 to 2010 was statistically significant with an increase of 78%.  The period from 2010 to 2017 showed a significant decrease of 39.8%.  An assessment of the annual differences indicated if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences. The results in figure \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to indicate which seasonal changes were contributing to the annual differences. Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to July.  Similarly, rows g-i show the seasonal averages and trend estimates from August to December.  The seasonal trend results suggested that the increase in annual averages from 2000 to 2010 was likely caused by an increase in chlorophyll values in the fall (Figure \@ref(fig:trnddat)h).  Seasonal changes from 2010 to 2017 showed that neither the January to July nor August to December changes were significant, suggesting an overall decrease in chlorophyll across all seasons may have recently occurred.

@@

chg1 <- anlz_perchg(mods = toplo, baseyr = 1991, testy = 2000)
chg2 <- anlz_perchg(mods = toplo, baseyr = 2001, testy = 2010)
chg3 <- anlz_perchg(mods = toplo, baseyr = 2011, testy = 2019)
chg <- bind_rows(chg1, chg2, chg3) %>% 
  mutate_if(is.numeric, round, 1) %>% 
  mutate(
    pval = p_ast(pval)
  )

@@

p1 <- show_perchg(mods = toplo, baseyr = 1991, testyr = 2000, ylab = ylab) +
  labs(
    subtitle = paste0('(a) 1991: ', chg[[1, 2]],' ug/L, 2000: ', chg[[1, 3]], 'ug/L\nChange: ', chg[[1,4]], '%, ', chg[[1,5]]), 
    title = NULL
  )
p2 <- show_perchg(mods = toplo, baseyr = 2000, testyr = 2010, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 1991: ', chg[[2, 2]],' ug/L, 2000: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
    title = NULL
  )
p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2019, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 1991: ', chg[[3, 2]],' ug/L, 2000: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
    title = NULL
  )
p4 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[1]],
    title = NULL
  )
p5 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[2]], 
    title = NULL
  )
p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) +  
  labs(
    subtitle = seas[[3]], 
    title = NULL
  )
p7 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[4]], 
    title = NULL
  )
p8 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[5]], 
    title = NULL
  )
p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[6]],
    title = NULL
  )

>>>>

For all results, model S was used with a sufficiently high number of knots in `num_knots_y` to evaluate chlorophyll trends across the monitoring stations in South San Francisco Bay.  This model was chosen because of the relatively quicker processing time to fit the model, while providing nearly identical explanatory power as compared to the other models (Table \@ref(tab:modstrtab)). Chlorophyll trends were generally well-explained across all stations (Table \@ref(tab:modprftab)), with an average R-squared value equal to `r aver2`% and ranging from `r minr2`% (station `r minst`) to `r maxr2`% (station `r maxst`). GAM predictions from north to south on the longitudinal axis provided different descriptions of annual and seasonal changes in chlorophyll (Figure \@ref(fig:prddat)). All models suggested an increase in chlorophyll in the period of record from 1990 to approximately 2005 to 2010, when predictions suggested a decreasing trend until the end of the record. Seasonally, all models indicated a larger spring peak in chlorophyll with more southern stations, whereas a subsequent peak in fall concentrations did not have any noticeable difference in magnitude by location (Figure \@ref(fig:prddat)).

@@

Examples of results provided by the trend tests are shown for station 34, with estimates of percent change between year pairs and seasonal trend results for January to June and July to December for the same year pairs (Figure \@ref(fig:trnddat)).  This figure illustrates the narrative descriptions that can be obtained for each test and the differences that can be observed depending on season.  Figures \@ref(fig:trnddat)a-c demonstrate that average chlorophyll at station 34 from 1990 to 2000 had a 268,4% increase, although the change is insignificant. Similarly, the change from 2000 to 2010 was an insignificant increase of 11%, followed by an insignificant decrease of 13.9% from 2010 to 2019.  Although these trends were insignificant, the results are based on a priori decisions of which years to compare.  Significant results may be observed for other year pairs.

An assessment of the annual differences indicates if changes overall were observed, but provides no information on seasonal shifts that may have contributed to these differences. Accordingly, figures \@ref(fig:trnddat)d-i show results from mixed-meta regression analysis to evaluate changes across the same year pairs but GAM results for different seasonal periods. Rows d-f show the seasonal averages and trend estimates for the same year periods in rows a-c, but only for months from January to June.  Similarly, rows g-i show the seasonal averages and trend estimates from July to December.  The seasonal trend results suggested that a significant increase in chlorophyll was observed from 1991 to 2000 but only for the January to June period (slope 0.57 ug/L chlorophyll change per year).  Similarly, A significant increase was observed from 2000 to 2010 but only for the July to December period (slope 0.47)  Finally, a significant decrease was observed from 2010 to 2019 but only for the July to December period (slope -0.29). As a result, an assessment of trend changes by season provides a more precise assessment of trends over time and accounts for full uncertainty in estimates across model results.

@@

chg1 <- anlz_perchg(mods = toplo, baseyr = 1991, testy = 2000)
chg2 <- anlz_perchg(mods = toplo, baseyr = 2000, testy = 2010)
chg3 <- anlz_perchg(mods = toplo, baseyr = 2010, testy = 2019)
chg <- bind_rows(chg1, chg2, chg3) %>% 
  mutate_if(is.numeric, round, 1) %>% 
  mutate(
    pval = p_ast(pval)
  )

@@

p1 <- show_perchg(mods = toplo, baseyr = 1991, testyr = 2000, ylab = ylab) +
  labs(
    subtitle = paste0('(a) 1991: ', chg[[1, 2]],' ug/L, 2000: ', chg[[1, 3]], 'ug/L\nChange: ', chg[[1,4]], '%, ', chg[[1,5]]), 
    title = NULL
  )
p2 <- show_perchg(mods = toplo, baseyr = 2000, testyr = 2010, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 2001: ', chg[[2, 2]],' ug/L, 2010: ', chg[[2, 3]], 'ug/L\nChange: ', chg[[2, 4]], '%, ', chg[[2, 5]]), 
    title = NULL
  )
p3 <- show_perchg(mods = toplo, baseyr = 2010, testyr = 2019, ylab = NULL) +
  labs(
    subtitle = paste0('(a) 2011: ', chg[[3, 2]],' ug/L, 2019: ', chg[[3, 3]], 'ug/L\nChange: ', chg[[3, 4]], '%, ', chg[[3, 5]]), 
    title = NULL
  )
p4 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[1]],
    title = NULL
  )
p5 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[2]], 
    title = NULL
  )
p6 <- show_avgseason(mods = toplo, doystr = 41, doyend = 213, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) +  
  labs(
    subtitle = seas[[3]], 
    title = NULL
  )
p7 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 1991, yrend = 2000, ylab = ylab) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[4]], 
    title = NULL
  )
p8 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2000, yrend = 2010, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[5]], 
    title = NULL
  )
p9 <- show_avgseason(mods = toplo, doystr = 213, doyend = 338, yrstr = 2010, yrend = 2019, ylab = NULL) +
  coord_cartesian(ylim = ylim) + 
  labs(
    subtitle = seas[[6]],
    title = NULL
  )
